---
title: "Longitudinal Analysis of Cognitive Age (ITTHACA)"
author: "Iñigo Tellaetxe Elorriaga"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: true
toc_float: true
knit: (
  function(inputFile, encoding) { 
    rmarkdown::render('cerebral_longitudinal_analysis.Rmd',
                      output_file = './rendered_R_notebooks/cerebral_longitudinal_analysis.html')
  })
---
  Load libraries

```{r}
# Clear workspace
rm(list = ls(all.names = TRUE))

# Load libraries
library(ggplot2); library(dplyr); library(sjPlot);library(metafor); library(ggpubr); library(tools); library(readxl); library(purrr); library(ggstatsplot)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
theme_set(theme_bw())

graph_th <- theme(
  text = element_text(family = "Times New Roman", size = 12),
  axis.text = element_text(family = "Times New Roman", size = 12),
  axis.title = element_text(family = "Times New Roman", size = 12),
  strip.text = element_text(family = "Times New Roman", size = 8),
  plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm"),
  axis.title.y = element_text(margin = margin(r = 2, unit = "mm")),
  axis.title.x = element_text(margin = margin(t = 2, unit = "mm"))
)
```


Load data and merge
```{r}
deltas_path <- "../analyses/longitudinal/cerebral_longitudinal/ageml/model_age/predicted_age.csv"

# Load deltas
deltas <- read.csv(deltas_path) %>%
  rename(delta = delta_all) %>%
  rename(X = id)

deltas <- deltas[c("X", "delta")]

# Load features
features_path <- "../data_for_ageml/longitudinal_data/cerebral_longitudinal/longitudinal_cerebral_features_with_record_id_SVH.csv"
features <- read.csv(features_path, stringsAsFactors=F, colClasses="character")

# Load the clinical file
clinical_path <- "../data_for_ageml/longitudinal_data/cerebral_longitudinal/longitudinal_clinical_file.csv"
clinical <- read.csv(clinical_path)

# Merge the three dataframes
deltas <- deltas %>%
  merge(features, by="X") %>%
  merge(clinical, by="X")

# Filter by only the record_id_SVHs that appear twice
deltas_filtered <- deltas %>%
  group_by(record_id_SVH) %>%
  filter(n() == 2)

# In deltas, make a column "group" taking the values of "CN", "CN_NonIntervened" and "CN_Intervened" based on those three columns when they are equal to 1
deltas_filtered <- deltas_filtered %>%
  mutate(group = case_when(
    CN == 1 ~ "CN",
    CN_NonIntervened == 1 ~ "CN_NonIntervened",
    CN_Intervened == 1 ~ "CN_Intervened"
  )) %>%
  # Also add a column named "timepoint" that equals 0 for "CN" in group, and 24 for the rest of values in the group column
  mutate(timepoint = case_when(
    group == "CN" ~ "t_0m",
    group == "CN_NonIntervened" ~ "t_12m",
    group == "CN_Intervened" ~ "t_12m"
  ))
# Make group and timepoint factors, and relevel them properly
deltas_filtered$group <- as.factor(deltas_filtered$group)
deltas_filtered$group <- relevel(deltas_filtered$group, ref="CN_NonIntervened")

deltas_filtered$timepoint <- as.factor(deltas_filtered$timepoint)
deltas_filtered$timepoint <- relevel(deltas_filtered$timepoint, ref="t_0m")
```


```{r}
table(deltas_filtered$group)
```


```{r}
# In deltas_filtered, for each "record_id_SVH", I want to assign it the group of the second timepoint
deltas_filtered <- deltas_filtered %>%
  group_by(record_id_SVH) %>%
  mutate(group = last(group)) %>%
  ungroup()
deltas_filtered$group <- relevel(deltas_filtered$group, ref="CN_NonIntervened")
```

```{r}
unique(deltas_filtered$group)
```

# Test progressions
Test the progression of the groups.
We use a Mixed-Model for Repeated Measures (MMRM) of this shape:
  $$y_{ij} = \beta_0 + \beta_1 * group + \beta_2 * timepoint + \beta_3 * group * timepoint + b_i + \epsilon_{ij}$$
  - Group: Intervention of Non-Intervention, fixed effect
- Timepoint: Either t=0 months or t=12 months, fixed effect
- $b_i$: Inter subject variability, random effect
- $\epsilon_{ij}$: Intra subject variability, random effect

We want to test the interaction term $\beta_3$, where $H_1: \mu_{\beta_3} \neq 0$

```{r}
library(lme4)
library(lmerTest)

model <- lmer(delta ~ group + timepoint + group:timepoint + (1 | record_id_SVH), data = deltas_filtered)
print(summary(model))
```


```{r}
qqnorm(resid(model))
qqline(resid(model))
```


```{r}
library(emmeans)

# Assuming you have your model already fitted:
# model <- lmer(delta ~ group + timepoint + group:timepoint + (1 | record_id_SVH), 
#               data = deltas_filtered)

# =============================================================================
# 1. VISUALIZE THE RESULTS
# =============================================================================

# Extract predicted values from the model
deltas_filtered$predicted <- predict(model)

# Plot 1: Spaghetti plot with trajectories
p1 <- ggplot(deltas_filtered, aes(x = timepoint, y = delta, group = record_id_SVH)) +
  geom_line(aes(color = group), alpha = 0.3) +
  geom_point(aes(color = group), alpha = 0.3) +
  # Add model predictions (population average)
  stat_summary(aes(group = group, color = group), 
               fun = mean, geom = "line", size = 1) +
  stat_summary(aes(group = group, color = group), 
               fun = mean, geom = "point", size = 5) +
  theme_minimal() +
  labs(title = "Individual Trajectories with Group Means",
       subtitle = "Thick lines = group averages",
       x = "Timepoint",
       y = "Delta",
       color = "Group") +
  theme(legend.position = "bottom")

print(p1)

# Plot 1: Spaghetti plot with model predictions
p2 <- ggplot(deltas_filtered, aes(x = timepoint, y = predicted, group = record_id_SVH)) +
  geom_line(aes(color = group), alpha = 0.3) +
  geom_point(aes(color = group), alpha = 0.3) +
  # Add model predictions (population average)
  stat_summary(aes(group = group, color = group), 
               fun = mean, geom = "line", size = 1) +
  stat_summary(aes(group = group, color = group), 
               fun = mean, geom = "point", size = 5) +
  theme_minimal() +
  labs(title = "Individual Trajectories with Group Means (Predictions)",
       subtitle = "Thick lines = group averages",
       x = "Timepoint",
       y = "Delta",
       color = "Group") +
  theme(legend.position = "bottom")

print(p2)
```

```{r}
# Plot 2: Box plots by group and timepoint
p3 <- ggplot(deltas_filtered, aes(x = timepoint, y = delta, fill = group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.2) +
  theme_minimal() +
  labs(title = "Distribution of Delta by Group and Timepoint",
       x = "Timepoint",
       y = "Delta",
       fill = "Group") +
  theme(legend.position = "bottom")

print(p3)
```

```{r}
# Plot 3: Interaction plot with confidence intervals
my_95CI <- function(data){mean_se(data, mult = 1.96)}

p4 <- ggplot(deltas_filtered, aes(x = timepoint, y = delta, color = group, group = group)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = my_95CI, geom = "errorbar", width = 0.1, size = 1) +
  theme_minimal() +
  labs(title = "Group Means with 95% Confidence Interval",
       subtitle = "",
       x = "Timepoint",
       y = "CAD",
       color = "Group") +
  theme(legend.position = "bottom")

print(p4)
ggsave(filename="CAD_longitudinal_trajectories.svg", plot=p4, width=6, height=4)
```

```{r}
# 1. Calculate the raw difference (Change)
library(tidyr)
plot_data <- deltas_filtered %>%
  filter(timepoint %in% c("t_0m", "t_12m")) %>%
  select(record_id_SVH, group, timepoint, delta) %>%
  pivot_wider(names_from = timepoint, values_from = delta) %>%
  mutate(abs_change = t_12m - t_0m) %>% # Just the subtraction
  filter(!is.na(abs_change))

# 2. Plot the Mean Change
plot_mean_change <- ggplot(plot_data, aes(x = group, y = abs_change, fill = group)) +
  stat_summary(fun = mean, geom = "bar", width = 0.6) +
  theme_minimal() +
  labs(
    title = "Mean Change in Cognitive Age Delta (t_12m - t_0m)",
    x = "Group",
    y = "Mean Change"
  ) +
  theme(legend.position = "none")

print(plot_mean_change)
ggsave(filename="CAD_mean_abs_change.svg", plot=plot_mean_change, width=6, height=4)
```

# Have the groups had a significant change in their delta over time?
```{r}
emm <- emmeans(model, ~ timepoint | group)
print(emm)
cat("====================================================\n")
print(contrast(emm, method = "pairwise", by = "group"))
```
The Intervened group has a statistically significant change in the delta, their delta got reduced.
The Non-Intervened group did not have a statistically significant change in their delta.

# Do the groups start or end with a different delta?
```{r}
emm <- emmeans(model, ~ group | timepoint)
print(emm)
cat("====================================================\n")
print(contrast(emm, method = "pairwise", by = "timepoint"))
```
At t=0m, the people that got Intervened in the future had a higher delta than the people that did not receive the intervention.
At t=12m, both groups have a similar mean delta.


```{r}
# =============================================================================
# 2. ESTIMATED MARGINAL MEANS
# =============================================================================

# Get estimated marginal means
emm <- emmeans(model, ~ group * timepoint)

print("=== ESTIMATED MARGINAL MEANS ===")
print(summary(emm))

# Pairwise comparisons between groups at each timepoint
print("\n=== PAIRWISE COMPARISONS: Groups within each Timepoint ===")
pairs_by_time <- pairs(emm, by = "timepoint")
print(pairs_by_time)

# Pairwise comparisons between timepoints within each group
print("\n=== PAIRWISE COMPARISONS: Timepoints within each Group ===")
pairs_by_group <- pairs(emm, by = "group")
print(pairs_by_group)

# # Test for differences in slopes (change over time)
# print("\n=== TEST OF SLOPES (Change from baseline to 12m) ===")
# emm_trends <- emtrends(model, ~ group, var = "timepoint")
# print(summary(emm_trends))
# print("\nCompare slopes between groups:")
# print(pairs(emm_trends))
# 
# # Visualize estimated marginal means
# emmip_plot <- emmip(model, group ~ timepoint, CIs = TRUE) +
#   theme_minimal() +
#   labs(title = "Estimated Marginal Means with 95% CI",
#        x = "Timepoint",
#        y = "Predicted Delta",
#        color = "Group") +
#   theme(legend.position = "bottom")
# 
# print(emmip_plot)

# =============================================================================
# 3. MODEL DIAGNOSTICS
# =============================================================================

# Set up plotting area for diagnostics
par(mfrow = c(2, 3))

# Diagnostic 1: Residuals vs Fitted
plot(fitted(model), residuals(model),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
lowess_line <- lowess(fitted(model), residuals(model))
lines(lowess_line, col = "blue", lwd = 2)

# Diagnostic 2: Q-Q plot of residuals
qqnorm(residuals(model), main = "Normal Q-Q Plot: Residuals")
qqline(residuals(model), col = "red", lwd = 2)

# Diagnostic 3: Scale-Location plot
plot(fitted(model), sqrt(abs(residuals(model))),
     xlab = "Fitted Values", ylab = "√|Residuals|",
     main = "Scale-Location")
lowess_line <- lowess(fitted(model), sqrt(abs(residuals(model))))
lines(lowess_line, col = "blue", lwd = 2)

# Diagnostic 4: Random effects Q-Q plot
random_effects <- ranef(model)$record_id_SVH[[1]]
qqnorm(random_effects, main = "Normal Q-Q Plot: Random Effects")
qqline(random_effects, col = "red", lwd = 2)

# Diagnostic 5: Histogram of residuals
hist(residuals(model), breaks = 30, main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue")

# Diagnostic 6: Residuals by group
boxplot(residuals(model) ~ deltas_filtered$group,
        main = "Residuals by Group",
        xlab = "Group", ylab = "Residuals",
        col = "lightgreen")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))  # Reset plotting area

# Statistical tests for assumptions
print("\n=== NORMALITY TESTS ===")
print("Shapiro-Wilk test for residuals:")
shapiro_resid <- shapiro.test(residuals(model))
print(shapiro_resid)

print("\nShapiro-Wilk test for random effects:")
shapiro_ranef <- shapiro.test(random_effects)
print(shapiro_ranef)

# Homogeneity of variance
print("\n=== HOMOGENEITY OF VARIANCE ===")
library(car)
levene_test <- leveneTest(residuals(model) ~ deltas_filtered$group)
print(levene_test)

# Influential observations
print("\n=== INFLUENTIAL OBSERVATIONS ===")
influence_data <- influence(model, obs = TRUE)
cook_d <- cooks.distance(influence_data)
influential <- which(cook_d > 4/length(cook_d))
print(paste("Number of influential observations (Cook's D > 4/n):", length(influential)))
if(length(influential) > 0) {
  print("Influential observation indices:")
  print(influential)
}
```







